---
title: "Sideways Volcano Hepatocellular Carcinoma"
author: "Manasvita Vashisth"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Identifying gene sets that depict a scaling relationship and studying their ability to predict survival"
output:
  html_document:
    df_print: paged
    number_sections: yes

---

```{r setup, include=FALSE, error=FALSE}
# Knit options

## Display code in final document
knitr::opts_chunk$set(echo = TRUE)

## Change working directory if data and Rmd files are in different directories
knitr::opts_knit$set(root.dir = '/Applications/UPenn/Summer2020')


# Package setup

## Install packages (initial run only)
#install.packages("survminer")
#install.packages("survival")

## Load packages
library(survminer)
library(survival)

# Load data
######reading the files into the program

rm(list=ls()) #remove any previous variables
df.input = read.table("LIHC_survival.txt", header = TRUE,sep="\t",stringsAsFactors = FALSE) #Phenotype data like survival years etc
df.map = read.table("Liver_Seq", header = FALSE,sep="\t",stringsAsFactors = FALSE,na.strings=c("", "NA")) #Matrix of genes vs patient id
df.map <- na.omit(df.map) #remove NA if any

```

# Physicochemical scaling concepts for correlated expression

## Polymer physics/fractal scaling

Collagens and lamins assemble into high-order fractals typical of structural polymers that impart architecture and properties – including stiffness scaling versus concentration with implications for gene-gene relationships. The concentration of such "mechanosensitive" proteins varies as a power law based on the stiffness of tissue (for example, bone being much stiffer than brain) [Swift et al](https://science.sciencemag.org/content/341/6149/1240104)
$$
E \sim A^\alpha \quad E \sim B^\beta 
\\ A \sim B^\frac{\beta}{\alpha}
$$

## Stoichiometric scaling

Collagen-I fibers are stoichiometric assemblies of COL1A1 and COL1A2 proteins, and a parallel increase of COL1A1 and COL1A2 transcripts (and protein) as a function of tissue stiffness can be useful in assessing data quality for normal tissues. Such stoichiometric scaling is linear and can reflect co-regulated mRNA upstream of protein assemblies, including expression from the same promoter (COL4A1 & COL4A2)

## Temporal scaling

Power laws also underlie highly time dependent biological processes such as cell cycle
$$
A \sim t^\alpha \quad B \sim t^\beta 
\\ A \sim B^\frac{\alpha}{\beta}
$$

# Processing the data pipeline

## Selecting Primary tumors

Narrowing down mRNA Seq entries to only primary tumor sites (mRNA seq files contain entries from different tissue type including adjacent univolved tissue)

```{r primary_tumor, include=FALSE, error=FALSE}
## narrowing down to only primary tumor sites
Trk1=df.map[2:dim(df.map)[1],2:3] #data frame storing mRNA seq values only
name=df.input[1] #patient ids from the survival file
namepro=df.map[1,2:dim(df.map)[2]] #patient ids from mrna seq file
patientid=substr(namepro[1],1,12) #pure patient id
k=1
for(i in 1:dim(namepro)[2])
{
  if(substr(namepro[1,i],14,15)=='01')
  {
    Trk1[,k]=as.double(df.map[2:dim(df.map)[1],i+1]) #mrna reads
    patientid[k]=substr(namepro[i],1,12)
    k=k+1
  }
}

```

## Matching RNA seq patient data to survival data

The mRNA Seq values for each patient are matched with the unique patient code identifying their overall survival
```{r matching_survival_mRNA, include=FALSE, error=FALSE}
##matching RNA seq patient data to survival data
Trk <- data.frame(matrix(ncol = 3, nrow = 2))
for(i in 1:length(patientid))
{
  for(j in 1:dim(df.input)[1])
  {
    if(patientid[i]==df.input$X_PATIENT[j] )
    {
      Trk[i,3]=df.input$OS[j]
      Trk[i,2]=df.input$OS.time[j]/365
      break
    }
  }
}


```

## Searching for gene of interest in the mRNA Seq data

```{r gene_search, include=FALSE, error=FALSE}
gene_of_interest="LMNB1"
iden=which(df.map$V1==gene_of_interest)-1

```

## Performing linear regression across all genes with respect to `r gene_of_interest`

Performing linear regression on the log values of the genes gives us the exponent of the power law as the slope of the fit. The $R^2$ determines the proportion of the variance in the dependent variable (log(Gene X)) that is predictable from the independent variable (log(LMNB1))
$$
[Gene X] \sim [LMNB1]^\beta
\\log(Gene X) \sim \beta log(LMNB1)
$$

```{r linear_regression, include=FALSE, error=FALSE}
## plotting the sideways volcano plot

chk=0.75*dim(Trk1)[2]  #remove any genes where 75% of the patients show zero reads
volc <- data.frame(matrix(ncol = 3, nrow = 2)) #storing the Rsq and slope values from all gene
k=1

for(i in 1:dim(Trk1)[1])
{
  if(sum(Trk1[i,]==0)<chk)
  {
    reg=as.data.frame(t(rbind(Trk1[i,],Trk1[iden,])))
    fit=lm(reg[,1]~reg[,2],data=reg)
    volc[k,1]=fit$coefficients[2]
    volc[k,2]=summary(fit)$r.squared
    volc[k,3]=i
    k=k+1
  }
}

# matching gene name after removing genes that have >75% zeros
gene=df.map[2:dim(df.map)[1],1] #all of the gene names
gene1=gene[volc[,3]]
volc1=cbind(volc,gene1)

# selecting strong scaling Rsq and Exponent
sbs=subset(volc1,volc1[,2]>0.5 & volc1[,1]>0.5)

write.table(sbs, file = "Liver_LMNB1_strong_scaling", append = FALSE, quote = TRUE, sep = "\t",
            eol = "\n", dec = ".", row.names = TRUE,
            col.names = TRUE, qmethod = c("escape", "double"),
            fileEncoding = "")  #change tumor name and gene name here

```

## Sideways Volcano Plot

```{r sideways_volcano,fig.width=10, fig.height=7,fig.cap="Sideways Volcano Plot"}
plot(volc[,2],volc[,1],pch=19,cex=0.8,xlab="Rsq",ylab="Exponent", cex.lab=1.5)
```

## Kaplan Meier

[Kaplan Meier](https://www.tandfonline.com/doi/abs/10.1080/01621459.1958.10501452) Plots enable us to observe significant differences in survival of patients based on their underlying mRNA Sequencing profile
```{r kaplan_meier, include=TRUE, fig.cap="Kaplan Meier Plot"}
Trk[,1]=as.double(t(Trk1[iden,]))
Trk[,4]=NULL
Trk[,4] <- Trk[,1]> median(Trk[,1])
fit <- survfit(Surv(Trk[,2],Trk[,3]) ~ Trk[,4],
               data = Trk)
ggsurvplot(fit, data = Trk, risk.table = FALSE,pval = TRUE,xlab = "Survival time in years",font.x=15,font.y=15,font.tickslab=15,legend.labs=c("low expressers","high expressers"))
```


## Evaluating patient survival 

Patients are divided into two cohorts (high expressers and low expressers) based on median expression on a particular gene. The median survival of both the cohorts is evaluated. This process is repeated for all genes.

```{r survival, include=FALSE, error=FALSE}
kmsig <- data.frame(matrix(ncol = 3, nrow = 2))
pro_name=df.map[1,1]
k=1
for (i in 1:dim(Trk1)[1])
{
  if(sum(Trk1[i,]==0)<chk)
  {
    Trk[,1]=as.double(t(Trk1[i,]))
    Trk[,4]=NULL
    Trk[,4] <- Trk[,1]> median(Trk[,1])
    fit <- survfit(Surv(Trk[,2],Trk[,3]) ~ Trk[,4],
                   data = Trk)
    
    kmsig[k,1]=surv_pvalue(fit)$pval
    kmsig[k,2]=surv_median(fit)$median[1]
    kmsig[k,3]=surv_median(fit)$median[2]
    pro_name[k]=df.map[i+1,1]
    k=k+1
  }
}

kmsig1=kmsig
kmsig2=cbind(kmsig1,pro_name)
kmsig2[,5]=kmsig2[,3]/kmsig2[,2]

write.table(kmsig2, file = "LIHC_Survival", append = FALSE, quote = TRUE, sep = " ",
            eol = "\n", na = "NA", dec = ".", row.names = TRUE,
            col.names = TRUE, qmethod = c("escape", "double"),
            fileEncoding = "") #Change Tumor Name here
sig=subset(kmsig2,volc1[,2]>0.5 & volc1[,1]>0.5)
sig1=subset(sig, sig[,1]<0.05)
per=dim(sig1)[1]/dim(sbs)[1]
```

## Survival Volcano Plot

The -log of ratio of median survivalof low expressers to high expressers is plotted on the y-axis. The x-axis plots the p-value of the Kaplan Meier Plot comparing survival of the two patients cohorts.

```{r survival_volcano, fig.width=10, fig.height=7, fig.cap="Sideways Survival Plot"}
plot(-log2(as.double(kmsig2[,1])),log2(as.double(kmsig2[,5])),pch = 19,cex=0.5,col="grey",xlab='-log2(p-value)',ylab='log2(median survival factor change)')
points(-log2(as.double(sig[,1])),log2(as.double(sig[,5])),pch = 19,cex=0.5,col="red")
points(c(-log2(0.05),-log2(0.05),-log2(0.05)),-1:1,pch = 3,cex=1)
```

Percentage of genes scaling strongly with `r gene_of_interest` that predict significantly poor survival=`r per*100`